On Optimality of Finite Element Integration

### 2016-01-23 TOMS_Paper/assembly.tex
Trying to improve again description of sharing elimination

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. We observe that, when multiplying two tensors, it often happens that the optimal scheduling strategy can be determined by comparing two possibilities. To illustrate the concept, consider the operation $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to $(a x^0_i + b x^1_i) (c x^0_i + d x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$. To minimize the operation count for sub-expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Applying generalized code motion, in presence of temporal sharing.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that applying Strategy~\ref{strategy:ii} would improve the operation count independently of the size of $L_i$, since the expression would be recast as $x^0_i (a + c) + x^1_i (b + d)$, thus exposing the two hoistable sub-expressions $(a + c)$ and $(b + d)$. On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$. In general, however, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions.
\begin{Def}[Structured expression]
\label{def:struct-expr}
An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in how to eliminate the spatial sharing induced by the symbols depending on $L$.
\end{Def}
The fundamental observation is the fact that a multilinear loop nest $L$ makes expressions structured with respect to $L$. In finite element integration loop nests, for example, a factorization of test and trial functions always results in exposing temporal sharing (Strategy~\ref{strategy:ii}). This will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality. Finally, it is remarkable that structured sub-expressions often characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).

In the following, we regard as \textit{structured} any sub-expression resulting from tensor operations for which there is no ambiguity in how to perform spatial sharing elimination -- implying that the optimal operation scheduling is to be determined out of Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. From the perspective of achieving quasi-optimality, it is fundamental to observe how multilinearity promotes structured expressions. : factorization of test and trial functions will always expose temporal sharing, \footnote{It is also remarkable that structured sub-expressions often characterize the \textit{entire} variational form. For instance, they systematically arise in the weak variational form of the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}.}. This is a consequence of multilinearity, which ensures that the factorization of any test or trial function will always expose some hoistable sub-expression (i.e., temporal sharing).

@@

\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i} to the multilinear operands. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. We will clarify that this directly descends from loop linearity and the structure inherent in tensor operations.}

@@

\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying Strategy~\ref{strategy:ii} would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
\textit{Note: this ensures the f-decreasing nature of the transformations at step (6)}.\\
\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost can be derived.}

\item If $\mathbb{S} = \mathbb{T}$, jump to step (6). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}. 
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii} such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
\textit{Note: this recasts the expression as $S (...S^1 + ... + ...S^n)$}.
\end{enumerate}
Go back to step (2).

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted sub-expression, apply the most profitable between strategies (i) and (ii). \\
\textit{Note: this pass speculatively assumes all sub-expressions to be structured. If the assumption does not hold, the result will generally be sub-optimal since only a subset of factorizations and code motion opportunities may be considered.}
\end{enumerate}
\end{Algo}

\begin{Rem*}
\normalfont
It is true that Algorithm~\ref{algo:sharing-elimination} aims for quasi-optimality, but clearly this does not mean that only those sub-expressions depending on the multilinear loop nest are considered for optimization. We emphasize once more that the enforcement of f-decreasing transformations (step 5) and the re-scheduling within outer loops (step 7) attempt to minimize the operation count of the whole loop nest. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}

>>>>

\subsubsection{Identification and exploitation of structure}
\label{sec:se-rln}
Finite element expressions can be seen as composition of operations between tensors. Often, the optimal implementation strategy for these operations is to be determined out of two alternatives. To illustrate the concept, consider $J^{-1} \nabla x \cdot J^{-1} \nabla x$, with $J^{-1}$ being the inverse Jacobian matrix for the change of coordinates and $x$ a generic two-dimensional vector. The tensor operation will reduce to $(a x^0_i + b x^1_i) (a x^0_i + b x^1_i) + ...$, in which $x^0_i$ and $x^1_i$ represent components of $x$ that depend on loop $L_i$. To minimize the operation count for sub-expressions of this kind, we have two options.
\begin{Strategy}
\label{strategy:i}
Eliminating temporal sharing through generalized code motion.
\end{Strategy}
\begin{Strategy}
\label{strategy:ii}
Eliminating spatial sharing first -- through product expansion and factorization -- and temporal sharing afterwards, again through generalized code motion.
\end{Strategy}
In the current example, we observe that, depending on the size of $L_i$, applying Strategy~\ref{strategy:ii} could reduce the operation count, since the expression would be recast as $x^0_i x^0_i a a + x^0_i x^1_i (ab + ab) + x^1_i x^1_i c c + ...$ and some hoistable sub-expressions would be exposed. On the other hand, Strategy~\ref{strategy:i} would have no effect as  $x$ only depends on a single loop, $L_i$. In general, the choice between the two strategies depends on multiple factors: the loop(s) size, the increase in operation count due to expansion (in Strategy~\ref{strategy:ii}), and the gain in code motion. Note that a second example of application of Strategy~\ref{strategy:ii} was provided in Figure~\ref{code:multi_loopnest}. These examples motivate the introduction of a particular class of expressions.
\begin{Def}[Structured expression]
\label{def:struct-expr}
An expression is \textit{structured} with respect to a set of loops $L$ if and only if there is no ambiguity in spatial sharing elimination for the symbols depending on $L$.
\end{Def}
The fundamental observation is that if a loop nest $L$ is multilinear, then all of its expressions will be structured with respect to $L$. In finite element integration loop nests, for example, a factorization of test and trial functions will always result in exposing temporal sharing (Strategy~\ref{strategy:ii}). This property will play a key role in Algorithm~\ref{algo:sharing-elimination}, which targets quasi-optimality. Finally, we observe that it is remarkable that structured sub-expressions sometimes characterize the \textit{entire} variational form (for instance, in the complex hyperelastic model analyzed in Section~\ref{sec:perf-results}).

@@

\item For each $P \in \mathbb{P}$ such that $|P| \leq |S_P|$, apply Strategy~\ref{strategy:i} to the multilinear operands. \\
\textit{Note: $|P|$ and $|S_P|$ represent the number of products in the innermost loop induced by $P$ if, respectively, Strategy~\ref{strategy:i} or Strategy~\ref{strategy:ii} were applied. This is a consequence of the expression being structured with respect to $[L_j, L_k]$.}

@@

\item Remove $e = (S^i, S^j) \in \mathbb{E}$ if applying Strategy~\ref{strategy:ii} would cause an increase in operation count (i.e., the elimination of temporal sharing after factorization of $S^i$ or $S^j$ would be outweighed by the expansion cost).\\
\textit{Note: this ensures the f-decreasing nature of the transformations at step (6)}.\\
\textit{Note: this pass can be accomplished by visiting the expression tree and collecting several kinds of information -- for example, the number of products wrapping an instance of $S^i$ or $S^j$, from which the expansion cost will be derived.}

\item If $\mathbb{S} = \mathbb{T}$, jump to step (7). Otherwise apply, in order, the following rewrite rules (\textit{precondition}: \textit{action}):
\begin{enumerate}[label=(\Alph*)]
\item $\lbrace T^1, ..., T^n \rbrace \subset \mathbb{T}$ adjacent to $S$, $n > 1$: apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T^1 + ... + ...T^n)$}.
\item Only one terminal $T$ adjacent to $S$: take an arbitrary non-terminal symbol $S^{nt}$ adjacent to $S$ and apply Strategy~\ref{strategy:ii}.\\
\textit{Note: this recasts the expression as $S (...T + ...S^{nt})$}. 
\item $\mathbb{T} = \emptyset$: take $S$ with highest degree and apply Strategy~\ref{strategy:ii} such that all symbols $\lbrace S^1, ..., S^n \rbrace$ adjacent to $S$ are grouped.\\
\textit{Note: this recasts the expression as $S (...S^1 + ... + ...S^n)$}.
\end{enumerate}
Go back to step (2).

\item Perform a depth-first visit of the expression tree and, for each yet unhandled or hoisted expression, apply the most profitable between Strategy~\ref{strategy:i} and Strategy~\ref{strategy:ii}. \\
\textit{Note: this pass speculatively assumes that all expressions are structured. If the assumption does not hold, the operation count within the reduction loop will generally be sub-optimal because only a subset of factorizations and code motion opportunities may eventually be considered.}
\end{enumerate}
\end{Algo}

\begin{Rem*}
\normalfont
Although the primary goal of Algorithm~\ref{algo:sharing-elimination} is operation count minimization within the multilinear loop nest, the enforcement of f-decreasing transformations (step 5) and the re-scheduling of sub-expressions within outer loops (step 7) also aim at optimizing the loop nest ``globally''. We will further elaborate this aspect in Section~\ref{sec:proof}.
\end{Rem*}
